Retrofitting Effect Handlers onto OCaml

### 2020-11-07 retro-concurrency.tex

Next we look at the cost of setting up an effect handler, but not |perform|ing
an effect. The aim is to quantify the cost of setting up and tearing down
effect handlers compared to a non-tail function call. To this end, we surround
the non-tail calls in the recursive micro benchmarks with an effect handler. We
also include a version of the same benchmarks implemented using concurrency
monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
the non-tail call and use an |MVar| to collect its result.

@@

Next we look at benchmarks that utilize non-local control flow. First, we
quantify the overhead individual operations in effect handling. Consider the
following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, ignoring 3 warm-up runs. We observed that
the sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
48ns, 10ns, 13ns and 19ns, respectively. For reference, the idle memory latency
for the local NUMA node is 93.2ns.

@@

\subsection{Measurements of overheads in our effects implementation}

We have implemented effects within OCaml, here we present timing measurements
for the various stages of stack creation (|a| to |b|), performing and handling
an effect (|b| to |c|), continuation resumption (|c| to |d|) and returning from
a computation and freeing the stack (|d| to |e|).

\begin{lstlisting}
let foo () =
 (* a *) match
 (* b *)   perform E
 (* d *) with effect E k ->
 (* c *)   continue k ()
 (* e *)
\end{lstlisting}

TODO:
\begin{itemize}
\item Aim: |R3| to provide some timing information as to the size of the overhead and where the overhead is.
\item use KC's measurement stubs as presented in WASM
\end{itemize}

>>>>

Next we look at the cost of setting up an effect handler, but not |perform|ing
an effect. The aim is to quantify the overhead of setting up and tearing down
effect handlers compared to a non-tail function call. To this end, we surround
the non-tail calls in the recursive micro benchmarks with an effect handler. We
also include a version of the same benchmarks implemented using concurrency
monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
the non-tail call and use an |MVar| to collect its result.

@@

Next we look at benchmarks that utilize non-local control flow. First, we
quantify the overhead individual operations in effect handling. Consider the
following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{48ns,
10ns, 13ns and 19ns}, respectively. For reference, the idle memory latency for
the local NUMA node is 93.2ns as measured using the Intel MLC tool~\cite{}.

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers, given any data structure (|'a t|)
and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
generator function (|val next : unit -> 'a option|) (|MC| version). The code is
available in the supplementary material.

We evaluate the performance of traversing a complete binary tree of depth 25
using this generator. For comparison, we also implemented a hand-written,
selective CPSed~\cite{}, defunctionalised~\cite{DefunctionalizationAtWork}
version (|cps|) and a concurrency monad (|monad|) version of the generator for
the tree. Both |cps| and |monad| versions are specialised to the binary tree,
and come with the usual caveats of not using the stack for function calls. We
observed that the |cps| version was the fastest thanks to specialisation and
hand optimisation. |MC| version was only 2.76$\times$ slower than |cps| while
being a generic solution, whereas the |monad| version was 8.69$\times$ slower
than |cps|.

@@
